Red Wine Quality Data Analysis by Suzanne Shoesmith
========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}


install.packages("GGally")
install.packages('dplyr')
install.packages("memisc")
install.packages("knitr")
install.packages("lmtest")
install.packages('gridExtra')
library(ggplot2)
library(dplyr)
library(memisc)
library(gridExtra)


```

```{r echo=FALSE, Load_the_Data}
# Load the Data
setwd('/Users/sshoesmith/Desktop/UdacityTerm2Project1')
pf <-read.csv('wineQualityReds.csv')

```

For my project I chose to analyze the data set containing wine 
quality ratings for red wine. This data set was one of the tidy data sets in
the suggested data sets in the Udacity link. The data set lists different 
phsyicochemical properties and corresponding wine quality ratings. The 
physicochemical properties include acidity, citric acid, sugar, alcohol content,
etc. The main variable of interest is quality. It will be 
interesting to determine if any of the other physicochemical
properties have an impact on the quality of the wine. 

# Univariate Plots Section

In the next few sections, I performed some basic exploration of the 
data (number of variables, number of columns and rows, and whether or not 
there are any null values). 


```{r echo=FALSE, Univariate_Plots}
names(pf)
summary(pf)
length(pf)
nrow(pf)
summary(pf$quality)


#check to see if there are any missing values: 
pf[!complete.cases(pf),]
str(pf)


#Plot to assess the distribution of wine quality

ggplot(aes(x=quality), data = pf) + geom_bar()
```

This value appears to be discrete, I'll determine the unique values 
in the variable price to determine the discrete values: 
```{r echo=FALSE}

table(pf$quality)

```
Based on the values obtained from the 
table command there are no ratings that are decimals, they are all whole numbers 
between 3 and 8. Also, based on the graph, it appears that the majority of wines 
are rated either a 5 or a 6. Also, there are no wines that were rated on either 
extreme of quality. 

The next variable that I will examine is Fixed Acidity. 

```{r echo=FALSE}
#Evaluating the remainder of the variables, making
#sure to use the min and max to appropriately set
#the scales

#Univariate Analysis Variable: Fixed Acidity

  
grid.arrange(
ggplot(pf, aes( x = 1, y = fixed.acidity)) + 
  geom_jitter(alpha = 0.1) + 
  geom_boxplot(alpha = 0.2, color = 'red'), 
  ggplot(pf, aes(x = fixed.acidity)) + 
geom_histogram(binwidth = 0.4) + scale_x_continuous() , ncol=2)

summary(pf$fixed.acidity)

```

I initially created a histogram with all the data, and then created a scatterplot
with a box plot in order to visually determine if there were any outliers that
I could remove the data set to get a better understanding of the variable 
fixed.acidity. 

I also summarized this variable Mean, Median, and quartile ranges in order to 
more accurately remove outliers. The interquartile range for this variable is:
9.20 - 7.10 = 2.1. 
An outlier would be 1.5 times this number in either direction. 
1.5 * 2.1 = 3.15
lower outliers: Q1 - 3.15
upper outliers: Q3 + 3.15

lower outlier limit: 7.10 - 3.15 = 3.95 (no lower outliers)
upper outlier limit: 9.20 + 3.15 = 12.35 (outliers are values greater than 
12.35, shown as the dots on the box plot)

I'll create another histogram below with the outliers removed. This will provide
a more accurate representation of this variable. 



```{r echo=FALSE}

ggplot(pf, aes(x = fixed.acidity)) + 
geom_histogram(binwidth = 0.3) + scale_x_continuous() +
  coord_cartesian(xlim = c(4.6,12.35))




```

With the new histogram with cartesian limits it is possible to see that 
although the data appears skewed to the right, it is less skewed than orginally
depicted in the histogram containing the outliers. 



```{r echo=FALSE}

grid.arrange(
ggplot(pf, aes( x = 1, y = volatile.acidity)) + 
  geom_jitter(alpha = 0.1) + 
  geom_boxplot(alpha = 0.2, color = 'red'), 
  ggplot(pf, aes(x = volatile.acidity)) + 
geom_histogram(binwidth = 0.035) + scale_x_continuous() , ncol=2)

summary(pf$volatile.acidity)

```

When looking at the plots for volatile acidity, there appear to be outliers
as inidicated by the red dots on the scatterplot. The summary of this variable 
is: 
Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 0.1200  0.3900  0.5200  0.5278  0.6400  1.5800
 
The interquartile range is 0.25, and 1.5x this number is 0.375
Since from the box plot it is obvious there are no lower outliers, 
I will use cartesian limits to remove upper outliers from the distribution. 
upper limit: 
0.64 + 0.375 = 1.015

The updated histogram is shown below: 


```{r echo=FALSE}
#updated histogram of volatile acidity using cartesian  limits removing the 
#upper outliers from the graph: 

ggplot(pf, aes(x = volatile.acidity)) + 
geom_histogram(binwidth = 0.05) + scale_x_continuous() +
  coord_cartesian(xlim = c(0.12,1.015))


```

After removing the outliers from the graph and adjusting the binwidth, 
the distribution of volatile acidity looks more like a normal distribution. 

The next variable I'll analyze is citric acid content: 

```{r echo=FALSE}


grid.arrange(
ggplot(pf, aes( x = 1, y = citric.acid)) + 
  geom_jitter(alpha = 0.1) + 
  geom_boxplot(alpha = 0.2, color = 'red'), 
  ggplot(pf, aes(x = citric.acid)) + 
geom_histogram(binwidth = 0.02) + scale_x_continuous() , ncol=2)

summary(pf$citric.acid)



```

Based on the above box plot, there  appears to be only 1 outlier in this
variable. The summary statistics are as follows: 

 Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  0.000   0.090   0.260   0.271   0.420   1.000 
  
The interquartile range is 0.33, and 1.5x this number is 0.495
Since from the box plot it is obvious there are no lower outliers, 
I will use cartesian limits to remove upper outliers from the distribution. 
upper limit: 
0.42 + 0.495 = 0.915

The updated histogram is shown below: 


```{r echo=FALSE}
#updated histogram with cartesian limits: 

ggplot(pf, aes(x = citric.acid)) + 
geom_histogram(binwidth = 0.02) + scale_x_continuous() +
  coord_cartesian(xlim = c(0.0,0.915))
```


Even after removing the outlier,  this distribution appears to be somewhat random. 
The next variable I will analyze is residual sugar: 


```{r echo=FALSE}

grid.arrange(
ggplot(pf, aes( x = 1, y = residual.sugar)) + 
  geom_jitter(alpha = 0.1) + 
  geom_boxplot(alpha = 0.2, color = 'red'), 
  ggplot(pf, aes(x = residual.sugar)) + 
geom_histogram(binwidth = 0.1) + scale_x_continuous() , ncol=2)

summary(pf$residual.sugar)
```

Based on the scatterplot and the initial histogram, there appear to be a large
number of high outliers, and the distribution is heavily skewed to the right 
with these outliers. 
The summary statistics are as follows: 

Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  0.900   1.900   2.200   2.539   2.600  15.500 

Removing the outliers and replotting: 

Interquartile range: 2.6 - 1.9 = 0.7
Upper outlier limit: 2.539 + (0.7*1.5) = 3.589
Lower outlier limit: 1.9 - (0.7*1.5) = 0.85

```{r echo=FALSE}
ggplot(pf, aes(x = residual.sugar)) + 
geom_histogram(binwidth = 0.1) + scale_x_continuous() +
  coord_cartesian(xlim = c(0.9,3.589))

```

After removing the outliers, the histogram of residual sugar appears to be more
normally distributed. 

The next variable to be analyzed is chlorides: 



```{r echo=FALSE}


grid.arrange(
ggplot(pf, aes( x = 1, y = chlorides)) + 
  geom_jitter(alpha = 0.1) + 
  geom_boxplot(alpha = 0.2, color = 'red'), 
  ggplot(pf, aes(x = chlorides)) + 
geom_histogram(binwidth = 0.01) + scale_x_continuous() , ncol=2)

summary(pf$chlorides)


```

Looking at the box plot, there appear to be outliers on both the upper and lower
edges of the distribution. 
The summary of the variable chlorides is as follows: 

  Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
0.01200 0.07000 0.07900 0.08747 0.09000 0.61100
  
The interqurtile range and outliers are: 
Interquartile range: 0.09 - 0.07 = 0.02
Upper outlier limit: 0.09 + (0.02*1.5) = 0.12
Lower outlier limit: 0.07 - (0.02*1.5) = 0.04

Replotting after removing the outliers from the plot: 

```{r echo=FALSE}
#replot after removing outliers using cartesian limits: 


ggplot(pf, aes(x = chlorides)) + 
geom_histogram(binwidth = 0.006) + scale_x_continuous() +
  coord_cartesian(xlim = c(0.04,0.12))

```

After adjusting the limits and the binwidth, the chlorides variable appears
to be normally distributed. 

The next variable to be analyzed is free sulfur dioxide: 


```{r echo=FALSE}
ggplot(aes(x = free.sulfur.dioxide), data = pf) +
  geom_histogram(binwidth = 2) +
  scale_x_continuous()

grid.arrange(
ggplot(pf, aes( x = 1, y = free.sulfur.dioxide)) + 
  geom_jitter(alpha = 0.1) + 
  geom_boxplot(alpha = 0.2, color = 'red'), 
  ggplot(pf, aes(x = free.sulfur.dioxide)) + 
geom_histogram(binwidth = 2) + scale_x_continuous() , ncol=2)

summary(pf$free.sulfur.dioxide)



```

Looking at the box plot it is clear that there are outliers in the variable 
free sulfur dioxide. The summary of the variable is as follows: 

 Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   1.00    7.00   14.00   15.87   21.00   72.00 
   
The interqurtile range and outliers are: 
Interquartile range: 21.00 - 7.00 = 14
Upper outlier limit: 21 + (14*1.5) = 42
Lower outlier limit: 7-(14*1.5) = -14


Replotting after removing the outliers from the plot: 
   
```{r echo=FALSE}
#replotting after removing outliers: 
ggplot(pf, aes(x = free.sulfur.dioxide)) + 
geom_histogram(binwidth = 1) + scale_x_continuous() +
  coord_cartesian(xlim = c(1,42))


```

After applying cartesian limits and adjusting the binwidth, the distribution
still appears to be skewed to the right. 

The next variable I"ll analyze is density: 


```{r echo=FALSE}

grid.arrange(
ggplot(pf, aes( x = 1, y = density)) + 
  geom_jitter(alpha = 0.1) + 
  geom_boxplot(alpha = 0.2, color = 'red'), 
  ggplot(pf, aes(x = density)) + 
geom_histogram(binwidth = 0.0005) + scale_x_continuous() , ncol=2)


summary(pf$density)



```

The distribution of density appears normal, but the box plot shows there are some
outliers, both on the lower and higher ends. The summary of the variable density
is as follows: 

Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 0.9901  0.9956  0.9968  0.9967  0.9978  1.0037
 
 The interqurtile range and outliers are: 
Interquartile range: 0.9978 - 0.9956 = 0.0022
Upper outlier limit: 0.9978 + 1.5(0.0022) = 1.0011
Lower outlier limit: 0.9956 - 1.5(0.0022) = 0.9923

replotting after applying cartesian limits: 

```{r echo=FALSE}
#replotting density after applying cartesian limits: 

ggplot(pf, aes(x = density)) + 
geom_histogram(binwidth = 0.0004) + scale_x_continuous() +
  coord_cartesian(xlim = c(0.9923,1.0011))
```

I replotted density using the cartesian limits and adjusted with binwidth. 
The data still appears normally distributed. After removing outliers, the tails
on either end are smaller. 

The next variable for analysis is pH: 



```{r echo=FALSE}


grid.arrange(
ggplot(pf, aes( x = 1, y = pH)) + 
  geom_jitter(alpha = 0.1) + 
  geom_boxplot(alpha = 0.2, color = 'red'), 
  ggplot(pf, aes(x = pH)) + 
geom_histogram(binwidth = 0.05) + scale_x_continuous() , ncol=2)


summary(pf$pH)




```

The distribution of pH in the dataset appears to be normally distributed with 
some outliers. The summary of the variable is as follows: 

Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  2.740   3.210   3.310   3.311   3.400   4.010


The interqurtile range and outliers are: 
Interquartile range: 3.4 - 3.210 = 0.19
Upper outlier limit: 3.4 + 1.5(0.19) = 3.685
Lower outlier limit: 3.210 - 1.5(0.19) = 2.925

replotting after applying cartesian limits: 

```{r echo=FALSE}
#replotting density after applying cartesian limits: 

ggplot(pf, aes(x = pH)) + 
geom_histogram(binwidth = 0.05) + scale_x_continuous() +
  coord_cartesian(xlim = c(2.925,3.685))
```

After removing the outliers and applying cartesian limits the pH data still 
appears normally distributed. 


```{r echo=FALSE}

grid.arrange(
ggplot(pf, aes( x = 1, y = sulphates)) + 
  geom_jitter(alpha = 0.1) + 
  geom_boxplot(alpha = 0.2, color = 'red'), 
  ggplot(pf, aes(x = sulphates)) + 
geom_histogram(binwidth = 0.03) + scale_x_continuous() , ncol=2)

summary(pf$sulphates)


```

The data for sulphates appears to be skewed to the right. Looking at the 
box plot there are outliers on the upper end of the distribution. 

The summary of the variable is as follows: 

Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
 0.3300  0.5500  0.6200  0.6581  0.7300  2.0000 
 
The interqurtile range and outliers are: 
Interquartile range: 0.73 - 0.55 = 0.18
Upper outlier limit: 0.73 + 1.5(0.18) = 1
Lower outlier limit: 0.55 - 1.5(0.18) = 0.28

replotting after applying cartesian limits: 

```{r echo=FALSE}

ggplot(pf, aes(x = sulphates)) + 
geom_histogram(binwidth = 0.03) + scale_x_continuous() +
  coord_cartesian(xlim = c(0.33,1))

```

In this case, even after removing the upper outliers, the distribution is still 
skewed to the right. 

The next variable to analyze is alcohol: 



```{r echo=FALSE}


grid.arrange(
ggplot(pf, aes( x = 1, y = alcohol)) + 
  geom_jitter(alpha = 0.1) + 
  geom_boxplot(alpha = 0.2, color = 'red'), 
  ggplot(pf, aes(x = alcohol)) + 
geom_histogram(binwidth = 0.1) + scale_x_continuous() , ncol=2)

summary(pf$alcohol)




```


The distribution of alcohol is skewed to the right. The summary of the variable 
is as follows: 

Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   8.40    9.50   10.20   10.42   11.10   14.90 
   
The interqurtile range and outliers are: 
Interquartile range: 11.1 - 9.5 = 1.6
Upper outlier limit: 11.1 + 1.5(1.6) = 13.5
Lower outlier limit: 9.5 - 1.5(1.6) = 7.1

replotting after applying cartesian limits to remove the upper outliers: 

```{r echo=FALSE}
ggplot(pf, aes(x = alcohol)) + 
geom_histogram(binwidth = 0.1) + scale_x_continuous() +
  coord_cartesian(xlim = c(8.4,13.5))
```

This variable still appears heavily skewed to the right. 



For all of the above plots, I chose binwidths that didn't overly smooth out the 
distribution, but not so small that it wasn't possible to observe
the shape of the distribution. 

There are a few observations I made about these single 
variable plots: 

- Most variables appear to be normally distributed or skewed to the right. I did 
not notice any variables that were skewed to the left. 

- Variables that appear to be slightly skewed to the right are
fixed acidity, volatile acidity, free sulfur dioxide, sulphates, and 
alcohol. That is most wines in 
the data set have low values for these variables. 

- Variables showing a long tail are alcohol 
and sulphates. 

- The histogram for citric acid appeared to be somewhat random, 
with most values at zero. 

Finally, looking at the distribution for alcohol content the distribution 
appears to be
somwhat skewed to the right, with the largest population around 9.5%. 

In addition to the univariate plots shown above, I created another variable, 
called alcohol_level, which is a categorical variable for the alcohol level. 
I divided the alcohol level into "low", "medium", and "high", using the levels: 

Alcohol percentage of 7.4 - 9.5: low
Alcohol percenrage of 9.5 - 11.1: medium
Alcohol percentage of 11.1 - 15: high

The plot below shows the distribution of this variable: 

```{r echo=FALSE}
#creating a new variable that will categorize alcohol level into 
#low, medium, and high for the purposes of graphing
pf$alcohol_level <- cut(pf$alcohol, breaks = c(7.40, 9.5, 11.10, 15), labels 
= c("low", "medium", "high"))
head(pf)

#creating the graph of the new variable: 

ggplot(pf, aes(alcohol_level)) + geom_bar() 
 


```

Based on the distribution of the alcohol level, most wines in the dataset fall 
into the medium alcohol level (that is, an alcohol % of 9.5 - 11.1). 


### What is the structure of your dataset?
The dataset contains 1599 rows and 13 variables. 
There are no null values. I also determined that the X column and the quality 
variables are integers while all other variables are numbers.

### What is/are the main feature(s) of interest in your dataset?
The main feature of interest in the dataset is quality rating. 


### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
The features in the dataset that would support the investigation are the 
physicochemical properties in the dataset. To name a few, these are alcohol, 
residual sugar, sulphates, etc.  

### Did you create any new variables from existing variables in the dataset?
The additional 
variables were alcohol_level (based on the quartile ranges). 

### Of the features you investigated, were there any unusual distributions? 
The distributions are descibed above. 
Did you perform any operations on the data to tidy, adjust, or change the form 
of the data? If so, why did you do this?
The data was already tidy, I didn't need to perform any additional operations. 


# Bivariate Plots Section

To start to gain an understanding of a possible correlation between 
the physicochemical properties and quality, I plotted a scatterplot matrix of 
all the variables in the dataset. I made a few interesting observations. 

```{r echo=FALSE}

setwd('/Users/sshoesmith/Desktop/UdacityTerm2Project1')
pf <-read.csv('wineQualityReds.csv')


#break it up into multiple plots to improve readability: 
pair1 <- data.frame(fixed_acidity = pf$fixed.acidity,
volatile_acidity = pf$volatile.acidity,
citric_acid = pf$citric.acid,
res_sugar = pf$residual.sugar,
chlorides = pf$chlorides,
quality = pf$quality)
GGally::ggpairs(pair1)  + theme_grey(base_size = 8)
        
        
pair2 <- data.frame(free_sulfur = pf$free.sulfur.dioxide,
total_sulfur = pf$total.sulfur.dioxide,
density = pf$density,
pH = pf$pH,
sulphates = pf$sulphates,
alcohol = pf$alcohol,
quality = pf$quality)
GGally::ggpairs(pair2)  + theme_grey(base_size = 8)
        
        
```

Another way to summarize the correlation coefficients is using ggcorr: 

```{r echo=FALSE}1
#summarizing coefficients using ggcor: 

setwd('/Users/sshoesmith/Desktop/UdacityTerm2Project1')
pf <-read.csv('wineQualityReds.csv')

GGally::ggcorr(pf, label = TRUE, label_size = 2, label_round = 2,
               hjust = 0.8, size = 3, name = "Correlation")  
                


```

The above ggcorr visualization more clearly summarizes the correlation 
coefficients for all the variables in the dataset. 


# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Using the scatterplot matrix, I looked at the variable
quality and looked at the correlation coefficients to 
determine where if at all there is a correlation, specifically with quality. 
The greatest correlation is a value of 0.476, which is the
correlation coefficient between the two variables alcohol (which refers to 
alcohol content) and quality. To further investigate this, I created a 
boxplot of alcohol content and quality, and use color to differentiate 
between the different alcohol levels. 


To further clarify, I created a box plot summarizing the alcohol content of 
different quality ratings. Alcohol content does seem to impact quality rating, 
but only when the alcohol content is above 10. It's also interesting to look at 
the median of the alcohol % for each quality level. 

```{r echo=FALSE}


#box plot of alcohol vs quality
pf$quality <- factor(pf$quality)
ggplot(aes(x = quality, y = alcohol),
  data = pf) + 
  geom_jitter(alpha = 0.3) + 
  geom_boxplot(alpha = 0.5, color = 'blue') + 
  stat_summary(fun.y = "mean",
               geom = "point",
               color = "red",
               shape = 8,
               size = 4)
  
#line showing the median of the alcohol rating vs. quality level
setwd('/Users/sshoesmith/Desktop/UdacityTerm2Project1')
pf <-read.csv('wineQualityReds.csv')

pf <- transform(pf,alcohol_sulphates = alcohol + sulphates)
head(pf)

pf2 <- pf %>%
  group_by(quality) %>%
  summarize(alcohol_sulphate_mean = mean(alcohol_sulphates),
  alcohol_sulphates_median = median(alcohol_sulphates),
  alcohol_median = median(alcohol),
  sulphates_median = median(sulphates),
  n = n())%>%
  ungroup()%>%
  arrange(quality)
  


  
```

As seen from the box plot, the quality rating seems to be increasing with the 
increase in the alcohol_median, but only above an alcohol media of 10 % and a 
quality rating of 5.  

Another possible correlation is quality with sulphates. The correlation 
coefficient is 0.251.  I plotted a box plot to look at this relationship.
I used limits to get a better view of the data without the outliers. When I 
decreased the scale of the y aixs to 95% of the values on the upper limit, the 
trend became a little more obvious. It appears that the between the wine quality 
values of 4 and 7, the amount of sulphates is higher. and then levels off 
between the quality ratings of 7 and 8. Just like in the previous set of graphs, 
I thought it would be interesting to calculate the median sulphate per quality 
rating and plot that vs. quality rating. 

```{r echo=FALSE}


#box plot of sulphates vs quality
pf$quality <- factor(pf$quality)
ggplot(aes(x = quality, y = sulphates),
  data = pf) + 
  geom_jitter(alpha = 0.3) + 
  ylim(0.25, quantile(pf$sulphates, 0.95)) +
  geom_boxplot(alpha = 0.5, color = 'blue') + 
  stat_summary(fun.y = "mean",
               geom = "point",
               color = "red",
               shape = 8,
               size = 4)
  
#calculating median of alcohol + sulphates per quality rating
setwd('/Users/sshoesmith/Desktop/UdacityTerm2Project1')
pf <-read.csv('wineQualityReds.csv')

pf <- transform(pf,alcohol_sulphates = alcohol + sulphates)
head(pf)

pf2 <- pf %>%
  group_by(quality) %>%
  summarize(alcohol_sulphate_mean = mean(alcohol_sulphates),
  alcohol_sulphates_median = median(alcohol_sulphates),
  alcohol_median = median(alcohol),
  sulphates_median = median(sulphates),
  n = n())%>%
  ungroup()%>%
  arrange(quality)
  


```

The box plot of sulphates vs. quality rating appear that as sulphates increase, 
the quality of the wine increases. 
Sulphates seem to be increasing with increasing wine quality 
rating. Unlike the alcohol content plots, this plot shows a more dramatic 
increase, followed by a levelling off at a quality rating of 7. 


Since I noticed that both alcohol and sulphates may be correlated with quality, 
I created an additional variable, the sum of alcohol content and sulphates, and 
plotted that against quality. I also plotted a summary of the median 
alcohol_sulphates variable vs. Quality for each quality rating. 

```{r echo=FALSE}

#creating a new variable called alcohol_sulphates
#that will sum up. alcohol and sulphates. 
setwd('/Users/sshoesmith/Desktop/UdacityTerm2Project1')
pf <-read.csv('wineQualityReds.csv')


pf <- transform(pf,alcohol_sulphates = alcohol + sulphates)
head(pf)

pf2 <- pf %>%
  group_by(quality) %>%
  summarize(alcohol_sulphate_mean = mean(alcohol_sulphates),
  alcohol_sulphates_median = median(alcohol_sulphates),
  alcohol_median = median(alcohol),
  sulphates_median = median(sulphates),
  n = n())%>%
  ungroup()%>%
  arrange(quality)
  

pf$quality <- factor(pf$quality)

ggplot(aes(x = factor(quality), y = alcohol_sulphates), data = pf) + 
  geom_jitter(alpha = 0.3) +
  geom_boxplot(alpha = 0.5, color = "blue") +
  stat_summary(fun.y = "mean",
               geom = "point",
               color = "red",
               shape = 8,
               size = 4)
  ylim(9.5, quantile(pf$alcohol_sulphates, 0.95))
  

```

Looking at the above box plot of alcohol + sulphates vs. quality, the trend 
appears very similar to alcohol vs. quality and sulphates vs. quality. There 
is a dip at the quality rating of 5, but overall there appears to be an increase 
in quality rating with an increase in the variable alcohol + sulphates. 

To further explore the relationship between alcohol content and quality, I made 
a bar graph summarizing the different alcohol levels, shaded by quality: 

```{r echo=FALSE}

pf$alcohol_level <- cut(pf$alcohol, breaks = c(7.40, 9.5, 11.10, 15), labels 
= c("low", "medium", "high"))
head(pf)

ggplot(pf, aes(quality)) + geom_bar(aes(fill = alcohol_level)) +
xlab('Quality Rating') +
ylab('Count') + 
ggtitle("Distribution of Quality Ratings")



```

It's interesting to note that the area in blue in the graph above (representing 
the "high" alcohol level, that is, a percentage greater than 11.10) seems to be 
increasing with the quality rating and the portion of the bar in green seems to 
be decreasing. This bar graph suggests that the proportion of wines that are 
higher in alcohol % is greater in wines with a higher quality rating when 
compared to a lower quality rating. To determine if this was the case, I 
calculated the proporting of high alcohol level wines in each quality group and 
plotted vs. quality: 

```{r echo=FALSE}
setwd('/Users/sshoesmith/Desktop/UdacityTerm2Project1')
pf <-read.csv('wineQualityReds.csv')

pf$alcohol_level <- cut(pf$alcohol, breaks = c(7.40, 9.5, 11.10, 15), labels 
= c("low", "medium", "high"))
head(pf)

pf3 <- pf %>%
  group_by(quality) %>%
  summarise(
  num_high_alc = sum((alcohol_level == 'high'), na.rm=TRUE),
  n(),
  mean_alcohol = mean(alcohol),
  proportion_high = num_high_alc/n())%>%
  ungroup()%>%
  arrange(quality)

head(pf3)

#plot
ggplot(aes(x = factor(quality), y = proportion_high),
  data = pf3) + geom_jitter() + xlab('Quality Rating') +
ylab('Proportion of Wines with Alcohol Percentage above 11.1') + 
ggtitle("Proportion of Wines with Alcohol Percentage above 11.1 vs. Quality 
Rating") + 
theme(text = element_text(size = 8))



```

Interestingly, there does appear to be an upward trend in the proportion of 
wines with an alcohol % of greater than 11.10 as the quality rating goes up. 
There is a dip at the quality rating of 5, but in general there appears to be an 
upward trend. What's interesting also is that at a quality rating of 8, almost 
80% of the wines are above an alcohol percentage of 11.10. 


### Did you observe any interesting relationships between the other features \

(not the main feature(s) of interest)?

I found some relationships the can be explained by the physicochemical 
properties. For example, acetic acid content decreases the pH. 
There was a correlation coefficient of -0.496 between alcohol content and 
density. Another relationship that makes sense is the relationship between pH 
and fixed acidity (pH decreases as fixed acidity increases, correlation -0.683). 
These variables are due to nature (that is, pH is defined by acidity).  

### What was the strongest relationship you found?

The strongest relationship that I found was between alcohol content and quality. 

# Multivariate Plots Section

Although based on what I've seen so far with the correlation coefficients and 
the visualizations, it appears that alcohol content has the greatest relationship 
to quality rating, with sulphates also appearing to be somewhat correlated with 
quality. However, the multivariate plots below look to further investigate any 
additional relationships between variables. 

In addition to alcohol content and sulphates, I'll plot variables that are 
related (based on the scatter plot matrix) such as pH and fixed acidity. 


```{r echo=FALSE}
pf$quality <- factor(pf$quality)

#scatter plot of alcohol vs. pH  (shaded by #quality)
ggplot(aes(x = pH, y = alcohol, color = factor(quality)), 
       data = pf) + 
  geom_point(alpha = 0.8, size = 1) +
  scale_color_brewer(type = 'seq', guide=guide_legend(title = "Quality")) + 
  geom_smooth(method = 'lm', se = FALSE, size = 1) + 
  theme_dark()
  
#scatter plot of pH vs. fixed acidity (shaded by quality)
ggplot(aes(x = fixed.acidity, y = pH, color = factor(quality)), 
       data = pf) + 
  geom_point(alpha = 0.8, size = 1) +
  scale_color_brewer(type = 'seq', guide=guide_legend(title = "Quality")) + 
  geom_smooth(method = 'lm', se = FALSE, size = 1)
  
  
#scatter plot of alcohol content vs. sulphates (shaded by quality)
ggplot(aes(x = alcohol, y = sulphates, color = factor(quality)), 
       data = pf) + 
  geom_point(alpha = 0.8, size = 1) + 
  ylim(0.25, 1.3) + 
  scale_color_brewer(type = 'seq', guide=guide_legend(title = "Quality")) + 
  geom_smooth(method = 'lm', se = FALSE, size = 1)


```

Looking at the first multivariate scatterplots in the above section (alcohol
vs pH shaded by quality with a linear regression model added) for quality
ratings 5, 6, 7, and 8, the linear model gets steeper with each increasing 
quality rated wine. In other words, there appears to be a stronger relationship 
of alcohol % to pH for higher quality wines when compared to lower quality
wines. 
 
For the plot of pH vs. fixed acidity, the relationship between pH and fixed 
acidity makes sense, since increasing acidity naturally decreased the pH. 
However, it's difficult to make out any patterns with regards to quality in 
this graph. 

The third multivariate plot reflects the same observation I made previously. 
That is, lower alcohol and lower sulphates appear to be correlated with lower 
quality, and higher sulphates and higher alcohol % seem to be correlated with 
higher quality ratings. That is, the darker lines corresponding to higher quality
wines are higher on the graph (corresponding to higher sulphates). It's less 
clear with alcohol content. 

As a follow up to the third graph, I am calculating the correlation coefficient 
of the alcohol+ sulphates variable with quality. 

```{r echo=FALSE}
setwd('/Users/sshoesmith/Desktop/UdacityTerm2Project1')
pf <-read.csv('wineQualityReds.csv')

pf <- transform(pf,alcohol_sulphates = alcohol + sulphates)


#calculate the correlation coefficient of the combined variable: 
cor.test(pf$alcohol_sulphates, pf$quality, method = 'pearson')
with(pf, cor.test(alcohol_sulphates, quality, method = 'pearson'))

#calculate the correlation coefficient with just the variable 
#alcohol
cor.test(pf$alcohol, pf$quality, method = 'pearson')
with(pf, cor.test(alcohol, quality, method = 'pearson'))

#calculate the correlation coefficient of quality with the variable
#sulphates
cor.test(pf$sulphates, pf$quality, method = 'pearson')
with(pf, cor.test(sulphates, quality, method = 'pearson'))

```

# Multivariate Analysis
Adding the sulphates and the alcohol variables 
together was an attempt to see an additive effect of both variables on quality, 
since looking at each individually there appeared to be a small correlation. 
This can also be done using a linear model. 


```{r echo=FALSE}
#linear model using alcohol content and sulphates

head(pf)

m1 <- lm(I(quality) ~ I(alcohol), data = pf)
m2 <- update(m1, ~ . + sulphates)

mtable(m1, m2)
```


### Were there any interesting or surprising interactions between features?
The interesting interaction was between alcohol and sulphates and 
the impact on quality. When added together these two variables resulted in a 
pearson's correlation coefficient of 0.5. It's surprising that sulphates would 
impact the correlation with quality. Without sulphates, the correlation 
coefficient was 0.476. 

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.
I created a linear model. The model included the variables alcohol and sulphates 
and quality. 
The model appears to be a weak model, with an R 
squared value of 0.227 without including sulphates and an R squared value of 
0.270 after including sulphates. 

------

# Final Plots and Summary



### Plot One
```{r echo=FALSE}
setwd('/Users/sshoesmith/Desktop/UdacityTerm2Project1')
pf <-read.csv('wineQualityReds.csv')



GGally::ggcorr(pf, label = TRUE, label_size = 2, label_round = 2,
               hjust = 0.8, size = 3, name = "Correlation")  


```

### Description One
The first graph that I chose is the ggcorr summary with all of the original
variables in the data set. I found this information most useful when 
starting the analysis. Looking at the correlation coefficients of quality with 
alcohol and sulphates, since they were the highest, helped in finding a few 
variables to start the analysis with. 


### Plot Two
```{r echo=FALSE}

#plot showing a bar graph color-filled by alcohol content
pf$alcohol_level <- cut(pf$alcohol, breaks = c(7.40, 9.5, 11.10, 15), labels 
= c("low", "medium", "high"))
head(pf)

ggplot(pf, aes(quality)) + geom_bar(aes(fill = alcohol_level)) +
xlab('Quality Rating') +
ylab('Count') + 
ggtitle("Bar Graph Distribution of Wine Quality Ratings Shaded by 
        Categorical % Alcohol Level") + 
guides(fill=guide_legend(title = "Categorical % Alcohol"))




```

### Description Two
I chose this as my second graph because I thought it shows very nicely how the 
proportion of blue (A high alcohol level as defined in this analysis) in each 
graph increases with increasing quality rating. At the same the proportion low 
alcohol wines seems to be decreasing with increasing quality. 


### Plot Three
```{r echo=FALSE}
#A plot of the proportion of high alcohol levels per #quality rating
#Overall disctribution of quality
setwd('/Users/sshoesmith/Desktop/UdacityTerm2Project1')
pf <-read.csv('wineQualityReds.csv')

#box plot of sulphates vs quality
pf$quality <- factor(pf$quality)
ggplot(aes(x = quality, y = sulphates),
  data = pf) + 
  geom_jitter(alpha = 0.3) + 
  ylim(0.25, quantile(pf$sulphates, 0.95)) +
  geom_boxplot(alpha = 0.5, color = 'blue') + 
  stat_summary(fun.y = "mean",
               geom = "point",
               color = "red",
               shape = 8,
               size = 4) +
      xlab('Wine Quality Rating') +
      ylab('Sulphate Content in Wine') + 
      ggtitle("Box Plot of Sulphate Content in Wine vs. 
        Wine Quality Rating")



```

### Description Three
I chose this graph for my final figure because it summarizes one of the
factors that I noticed had the strongest relationship to quality rating 
(sulphates). From this box plot, it's easy to see a gradual increase in wine 
quality rating with increasing sulphate concentration. 
------

# Reflection


I found the project very interesting, and enjoyed working through the analysis 
and finding which physicochemical proporties could impact the quality rating. I 
found that alcohol level (and possibly sulphate content) seemed to be related to 
quality rating. 
The difficulty came in explaining the dip in the alcohol content at the quality 
rating of 5. This could be due to several factors, but one that I can think of 
is the error/subjectiveness of the assignment of the quality rating to the 
different wines. The process is manual, and based on sensory response from the 
judges. Also, perhaps it is easier for a wine taster to discern a very good wine 
from a very bad wine, but the values in the middle may become more difficult 
to assign. 

Another challenge comes from the limited number of ratings. The range of quality 
ratings span only from 3 to 8. This gives us a smaller space to work in with 
regards to finding any trends in the data. 

The matrix scatterplot helped in terms of finding a place to start in terms of 
which variables could be related to wine quality. Perhaps I would've started 
with acidity instead of sulphates. Using the scatterplot I was able to narrow 
in on a couple of variables. It surprised me that sulphates were mildly related 
to rating, I wouln't have guessed that. 

Future work that can be done with this dataset could entail expanding the 
dataset to include a wider range of quality ratings. This could possibly clarify 
any possible relationships between physicochemical proporties and wine quality. 



